Highly Available Transactions: Virtues and Limitations

### 2013-03-28 writeup/evaluation.tex
Moar writing

\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient result describing exactly which
application logics can be achieved with high availability and to make
these requirements more concrete, we consider an example application:
the TPC-C benchmark. Our goal is \textit{not} to provide a performance
analysis, as any HAT configuration will be non-compliant. Rather, we
examine the five queries contained in the benchmarks---ostensibly
representative of a standard transaction processing application---to
understand which are achievable in a HAT context. TPC-C consists of
five transactions, capturing operation of a real-world warehouse,
including sales, payments, and deliveries. We analyze each kind of
transaction and its interactions below.

\vspace{.5em}\noindent\textit{New-Order} places an order for a
variable quantity of data items, updating warehouse stock as
needed. The transaction selects a sales district, assigns the order an
ID number, adjusts the remaining warehouse stock, and writes a
placeholder entry for the pending order. There are two mutations we
must be consider in detail.

First, assigning an order ID number is challenging. We can easily
assign a \textit{unique} order ID, by, say, concatenating the customer
ID with a timestamp and server ID. However, the TPC-C specification
requires that order numbers are assigned sequentially, which we cannot
guarantee in HATs because assigning sequential IDs requires preventing
Lost Update. Accordingly, HATs can allow placement of new orders with
unique identifiers but cannot provide compliant TPC-C execution due to
the requirement for a non-monotonic counter increase.

Second, decrementing inventory counts is, in general, a potentially
risky operation because, without coordination, inventory counts may
become negative. However, TPC-C automatically ``restocks'' inventory
counts (by 91 items) when there are 10 or fewer items
remaining. Accordingly, even in the presence of concurrent new orders
to the same item, the stock will never fall below zero. However, an
$R$ replica HAT system might end up with $91(R-1)$ more stock than in
a single-site implementation.

\vspace{.5em}\noindent\textit{Payment} updates running balances for
warehouses, districts, and customer records as well as providing an
audit trail. All balance increase operations can be made commutative,
so the transaction can execute under HATs.

\vspace{.5em}\noindent\textit{Order-Status} is a read-only transaction
and can be safely executed under HATs. The TPC-C specification does
not require recency bounds but, in practice, they may be desirable. If
a client is sticky with a set of replicas, she will read her own
writes, which may be sufficient.

\vspace{.5em}\noindent\textit{Delivery} represents the fulfillment of
a pending order created by a New Order transaction: it deletes the
order from the pending list, updates the customer's balance, updates
the order's carrier ID and delivery time, and updates the customer
balance. The challenge in delivery comes in deletion, which is
non-monotonic: an item might be removed twice from the pending
table. If delivery can be made idempotent---that is, if multiple
Delivery transactions have the same effect as one---then multiple
clients can execute the same Deliveries without fatal consequences.

In practice, given that a physical shipment is likely only delivered
once, it is unlikely that multiple agents will attempt a Delivery
transaction for the same order simultaneously. However, in this case,
we have simply moved the non-monotonicity to the real world: the
carrier that picks up the package for an order must ensure that she is
the only carrier who has done so. While we can program around the
other operations in the order, HATs cannot--without the help of
business-specific processes--serve Delivery transactions. That said,
New-Orders may be placed with high availability.

\vspace{.5em}\noindent\textit{Stock-Level} is another read-only
transaction and can be executed without any non-HAT guarantees, with
the same caveats regarding recency.

\vspace{.5em} Throughout execution, TPC-C also requires the
maintainence of several integrity constraints. For example,
Consistency Condition 1 (3.3.2.1) requires that each warehouse's sales
count must reflect the sum of its subordinate sales districts. This
integrity constraint spans two tables but, given the ability to update
both tables atomically via transactional atomicity, can be easily
maintained. Consistency Conditions 4 through 12 (3.3.2.4-12) can
similarly be satisfied by applying updates atomically across
tables. Consistency Conditions 2 and 3 (3.3.2.2-3) concern order ID
assignment and are more problematic, as we have discussed.

>>>>

\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient
result describing exactly which application logics can be achieved
with high availability and to make these requirements more concrete,
we consider an example application: the TPC-C benchmark. Our goal is
\textit{not} to provide a performance analysis, as any HAT
configuration will be non-compliant. Rather, we examine the five
queries contained in the benchmarks---ostensibly representative of a
standard transaction processing application---to understand which are
achievable in a HAT context. Surprisingly, four of five can be
executed with HATs, while the fifth potentially requires unavailablity.

TPC-C consists of five transactions, capturing operation of a
real-world warehouse, including sales, payments, and deliveries. Two
transactions---Order-Status and Stock-Level---are read-only and can be
executed safely under HATs. Clients may read stale data, but this does
not violate TPC-C requirements and clients will read their writes if
they are sticky-available. Another transaction type, Payment, updates
running balances for warehouses, districts, and customer records as
well as providing an audit trail. The transaction is increment-only,
so all balance increase operations can commute, and TA allows the
maintainence of any foreign-key constraints.

While three out of five transactions have been easily achievable with
HATs, the remaining two transactions---New-Order and Delivery---are
not so simple. The New-Order transaction places an order for a
variable quantity of data items, updating warehouse stock as
needed. It selects a sales district, assigns the order an ID number,
adjusts the remaining warehouse stock, and writes a placeholder entry
for the pending order. The Delivery transaction represents the
fulfillment of a New-Order: it deletes the order from the pending
list, updates the customer's balance, updates the order's carrier ID
and delivery time, and updates the customer balance.

There are three challenges that only allow HATs to place non-compliant
New-Orders and to fail on Payments. First, we must assign each new
order an ID number. We can assign a unique number by, say,
concatenating the client ID and a timestamp. However, the TPC-C
specification requires order numbers to be assigned sequentially,
which requires Lost Update. Accordingly, HATs cannot provide compliant
TPC-C execution but can maintain uniqueness constraints. Second, the
New-Order transaction decrements inventory counts: what if the count
becomes negative? Fortunately, TPC-C New-Order restocks the inventory
count (increments by 91) if it would become negative as the result of
placing an order. This means that, even in the presence of concurrent
New-Orders, an item's stock will never fall below zero. This is TPC-C
compliant, but a HAT system might end up with substantially more stock
than in a single-site implementation. Third, delivery events are
non-monotonic---deleting a pending New-Order transaction---and should
be idempotent in order to avoid billing a customer twice; this implies
a need to prevent Lost Update. We can avoid this issue by moving the
non-monotonicity to the real world---the carrier that picks up the
package for an order can ensure that she is the only carrier who has
done so---but cannot provide a correct execution with HATs alone.

Throughout execution, TPC-C also requires the maintainence of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update both tables atomically via
transactional atomicity, can be easily maintained. Consistency
Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
applying updates atomically across tables. Consistency Conditions 2
and 3 (3.3.2.2-3) concern order ID assignment and are more
problematic, as we have discussed.

### 2013-03-28 writeup/hat-analysis.tex
Moar writing

To begin, Adya's \textit{Read Uncommitted} isolation requires that
transactions are totally ordered and that writes within transactions
are ordered consistently with this order (prohibiting ``Dirty
Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
set of data items, then the final database state cannot contain the
``earlier'' transaction's writes to the data items. For example, in
the below example, $T_3$ should eventually only read $a=b=1$ or
$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\\[-2em]
\end{align*}
The difference between this total order on transactions and the total
order required by serializability is that the Read Uncommitted total
order is completely arbitrary (i.e., an order must exist), whereas a
serializable total order must be equivalent to a serial
execution. Read Uncommitted is similar to the distributed systems
concept of replica convergence, in which all replicas for a data item
must end up with the same data item. This convergence is usually
achieved by picking a ``winning'' update, requiring a means of totally
ordering updates to each item; one can view Read Uncommitted as a
cross-item convergence property. It is easily achieved via applying
the same logical timestamp to each update in a transaction.

@@

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are promised by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatabable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$'s must see $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read $\bot$ for each data item. However, when coupled with
additional properties, like transactional atomicity, it is stronger:
the repeated reads must obey additional ordering constraints. In fact,
distributed systems describe a ``consistent snapshot'' across a set of
related events as a \textit{cut} across a set of participants or, in
our case, data items. Accordingly, to capture the notion that the
transaction should read from a (non-changing) consistent cut over data
items (and to disambiguate from the aforementioned stronger Repeatable
Read properties), we call ANSI SQL Repeatable Read (preventing P2)
\textbf{Cut Isolation}.It is possible to satisfy cut isolation with
high availability by caching appropriate versions for a transaction's
reads or, alternatively, using multi-versioning on the server and
ensuring that each transaction's successive reads return the same
version of each data item.

@@

Transactional atomicity is core to ACID guarantees. Although, at least
by the acronym, it is not an ``isolation'' property, transactional
atomicity restricts transactions' ability to view the effects of
partially completed transactions. Under transactional atomicity,
within each transaction, either all effects of another transaction are
observed, or none are (equivalently, once some of the effects of a
transaction are observed, all effects are observed). This is a
strictly stronger guarantee than Read Committed isolation: in Read
Committed, we can read a subset of committed writes whereas, with
transactional atomicity, we must read all of them. For example, $T_2$
must observe $b=c=1$ (or later versions for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$ or $a=1$.

Perhaps perplexingly, discussions of transactional atomicity are
absent from existing discussions of weak isolation levels. This is
perhaps again due to the single-node context in which prior work was
developed: on a single server (or a fully replicated database),
atomicity is near-trivial to achieve via lightweight locking and/or
local concurrency control over data items. (In this spirit, multi-item
atomicity is common in so-called ``entity groups,'' which logically
represent groups of co-located items~\cite{entitygroup}.) In contrast,
in a distributed environment, atomicity over arbitrary groups of items
is considerably more difficult to achieve with high availability. The
key challenge is that different servers responsible for subsets of a
transaction's updates must coordinate regarding when to reveal their
writes to a transaction---without violating availability due to, say,
locking, or a missing transactional write. Together with cut
isolation, transactional atomicity prevents Read Skew anomalies
(Berenson et al.'s A5A~\cite{ansicritique}).

Transactional atomicity is achievable in a HAT environment. To ensure
transactional atomicity, servers need to know when all of a
transaction's updates are present on their respective
replicas. However, replicas do not need to agree on when to reveal a
new value to clients. In distributed systems terminology,
transactional atomicity only requires reliable broadcast (with some
additional metadata for clients). We omit a full discussion of the
algorithm and instead refer the reader to the extended version of this
paper~\cite{hat-tr}.

@@

So far, our models have not considered what happens across
transactions: we have not guaranteed any ordering or continuity
between transactions (other than that there exists some arbitrary
ordering). However, in the distributed systems context, many useful
\textit{safety} guarantees span multiple application actions.
\textit{Session guarantees} are used to describe guarantees across
groups of operations (here, transactions) issued by the same logical
user~\cite{sessionguarantees}. Informally, a \textit{session} is an
application context that consists of logical state that should persist
between operations. For example, users accessing a web site may want
to experience a notion of ``forward progress'' for the duration that
they are logged in; all transactions submitted between ``log in'' and
``log out'' operations might form a session. We can define sessions at
several granularities (e.g., per-transaction, per-client, per-time unit)
but, for simplicity, will refer to client-level guarantees in the
discussion.

@@

\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} requires that
a client's subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

\vspace{.5em}\noindent\textit{\textbf{Monotonic writes}} requires
that transactions from each individual client be serialized. Any order
on transactions should also respect the order in which each session
submitted transactions.

@@

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until their respective dependencies are fulfilled on
all replicas. In the case of Monotonic Reads, servers should only
reveal new writes to a data item when all other servers responsible
for the item have also seen them. In the case of Monotonic Writes and
Writes Follow Reads, each new write should become visible only when
each write's dependencies are visible everywhere. This mechanism
effectively requires that all clients read from a globally agreed upon
lower bound on the versions written to the system. This is highly
available as a client will never block due to inability to find a
server with a sufficiently up-to-date version of a data item. However,
it does not imply that transactions will read their own writes or, in
the presence of partitions, make forward progress through the version
history. The problem is that, if a server becomes partitioned, under
the highly available model, we must handle the possibility that an
unfortunate client will be forced to issue her next requests against
the partitioned server!

@@

Read your writes is unavailable in a highly available system. Consider
a client that executes the following two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server which is partitioned from the rest of the cluster, then the
client will not be able to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes. Accordingly, read your writes requires
stickiness, and, because causal consistency requires read your writes,
causal consistency also requires stickiness.

However, read-your-writes can be accomplished via client-side caching
and alternatively comes ``for free'' when requests are routed to a
sticky set of servers, while causality and PRAM can be accomplished
via sticky versions of the prior algorithms.

@@

\noindent{\textbf{Consistency}} A HAT system can make limited
consistency guarantees. It can execute commutative and logically
monotonic~\cite{calm} operations without the risk of invalidating
(also monotonic) application-level integrity constraints. Our goal in
this paper is not to sketch the entire space of consistency models
that are achievable (see Section~\ref{sec:futurework}, however we
specifically evaluate TPC-C transaction semantics under HAT
consistency guarantees in Section~\ref{sec:evaluation}.

@@

\vspace{.5em}\noindent{\textbf{Convergence}} Finally, to
require that the system propagates writes between replicas, we can
require convergence, or eventual consistency for each data item: in
the absence of new mutations to a data item, in the absence of
partitions, all servers eventually agree on the value for each
item. This is typically accomplished by any number of anti-entropy
protocols, which periodically update neighboring servers with the
latest value for each data item. This convergence property should be
considered mandatory given the spirit of existing literature on
isolation protocols: in all models we have encountered, the database
should behave as if there is some total order over
transactions. Indeed, non-convergent systems are remarkably difficult
to reason about.

>>>>

To begin, Adya's \textit{Read Uncommitted} isolation requires that
transactions are totally ordered and that writes within transactions
are ordered consistently with this order (prohibiting ``Dirty
Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
set of data items, then the final database state cannot contain the
``earlier'' transaction's writes to the data items. For example, in
the below example, $T_3$ should eventually only read $a=b=1$ or
$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\\[-2em]
\end{align*}
The difference between this total order on transactions and the total
order required by serializability is that the Read Uncommitted total
order is completely arbitrary (i.e., an order must exist), whereas a
serializable total order must be equivalent to a serial execution. It
is easily achieved via applying the same logical timestamp to each
update in a transaction.

@@

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are promised by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$'s must see $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the observable value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read $\bot$ for each data item. However, when coupled with additional
properties, like transactional atomicity, it is stronger: the repeated
reads must obey additional ordering constraints. In fact, distributed
systems describe a ``consistent snapshot'' across a set of related
events as a \textit{cut} across a set of participants or, in our case,
data items. Accordingly, to capture the notion that the transaction
should read from a (non-changing) consistent cut over data items (and
to disambiguate from the aforementioned stronger Repeatable Read
properties), we call ANSI SQL Repeatable Read (preventing P2)
\textbf{Cut Isolation}. It is possible to satisfy cut isolation with
high availability by caching appropriate versions for a transaction's
reads or, alternatively, using multi-versioning on the server and
ensuring that each transaction's successive reads return the same
version of each data item.

@@

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the acronym, it is not an ``isolation'' property,
transactional atomicity restricts transactions' ability to view the
effects of partially completed transactions. Under transactional
atomicity, within each transaction, either all effects of another
transaction are observed, or none are (equivalently, once some of the
effects of a transaction are observed, all effects are observed). This
is a strictly stronger guarantee than Read Committed isolation: in
Read Committed, we can read a subset of committed writes whereas, with
transactional atomicity, we must read all of them. Together with cut
isolation, transactional atomicity prevents Read Skew anomalies
(Berenson et al.'s A5A~\cite{ansicritique}). As an example of
TA, $T_2$ must observe $b=c=1$ (or later versions
for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$ or $a=1$.

Perhaps perplexingly, discussions of TA are absent from existing
discussions of weak ACID. This is perhaps again due to the single-node
context in which prior work was developed: on a single server (or a
fully replicated database), TA is near-trivial to achieve via
lightweight locking and/or local concurrency control over data
items. In contrast, in a distributed environment, TA over arbitrary
groups of non-colocated items is considerably more difficult to
achieve with high availability: servers need to know when all of a
transaction's updates are present on their respective
replicas. However, replicas do not need to agree on when to reveal a
new value to clients, which would require consensus; accordingly, TA
only requires reliable broadcast (with some additional metadata for
clients) and is achievable in a HAT system. We omit a full discussion
of the algorithm and instead refer the reader to the extended version
of this paper~\cite{hat-tr}.

@@

So far, our models have not considered what happens across
transactions: we have not guaranteed any ordering or continuity
between transactions (other than that there exists some arbitrary
ordering). However, in the distributed systems context, many useful
\textit{safety} guarantees span multiple application actions.
\textit{Session guarantees} are used to describe guarantees across
groups of operations (here, transactions) issued by the same logical
user~\cite{sessionguarantees}. A \textit{session} is an application
context that consists of logical state that should persist between
operations. For example, all of a user's transactions submitted
between ``log in'' and ``log out'' operations might form a session.

@@

\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} requires that,
with a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

\vspace{.5em}\noindent\textit{\textbf{Monotonic writes}} requires that
each session's writes be serialized. Any order on transactions should
also respect the order in which each session submitted transactions.

@@

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until their respective dependencies are fulfilled on
all replicas. This mechanism effectively requires that all clients
read from a globally agreed upon lower bound on the versions written
to the system. This is highly available as a client will never block
due to inability to find a server with a sufficiently up-to-date
version of a data item. However, it does not imply that transactions
will read their own writes or, in the presence of partitions, make
forward progress through the version history. The problem is that, if
a server becomes partitioned, under the highly available model, we
must handle the possibility that an unfortunate client will be forced
to issue her next requests against the partitioned server!

@@

Read your writes is unavailable in a highly available system. Consider
a client that executes the following two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server which is partitioned from the rest of the cluster, then the
client will not be able to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes. Accordingly, read your writes requires
stickiness, and, because causal consistency requires read your writes,
PRAM and causal consistency also require stickiness. Read-your-writes
is ``free'' in a sticky system, while the remaining causality and PRAM
guarantees can be accomplished with the prior algorithms.

@@

\noindent{\textbf{Consistency}} A HAT system can make limited
consistency guarantees. It can often execute commutative and logically
monotonic~\cite{calm} operations without the risk of invalidating
(also monotonic) application-level integrity constraints. Our goal in
this paper is not to sketch the entire space of consistency models
that are achievable (see Section~\ref{sec:futurework}, however we
specifically evaluate TPC-C transaction semantics under HAT
consistency guarantees in Section~\ref{sec:evaluation}.

@@

\vspace{.5em}\noindent{\textbf{Convergence}} To require that the
system propagates writes between replicas, we can require convergence,
or eventual consistency for each data item: in the absence of new
mutations to a data item, in the absence of partitions, all servers
eventually agree on the value for each item. This is typically
accomplished by any number of anti-entropy protocols, which
periodically update neighboring servers with the latest value for each
data item. Establishing a final value is related to determining a
total order on transaction updates, as in Read Uncommitted.
